---
title: "p8105_hw2_jt3174"
author: "Jingyi"
date: "9/25/2019"
output: github_document
---

## **Problem 0**

This “problem” focuses on structure of your submission, especially the use git and GitHub for reproducibility, R Projects to organize your work, R Markdown to write reproducible reports, relative paths to load data from local files, and reasonable naming structures for your files.

To that end:

* create a public GitHub repo + local R Project; we suggest naming this repo / directory p8105_hw2_YOURUNI (e.g. p8105_hw2_ajg2202 for Jeff), but that’s not required
* create a single .Rmd file named p8105_hw2_YOURUNI.Rmd that renders to **github_document**
* create a subdirectory to store the local data files used in Problems 1 and 2, and use **relative paths** to access these data files
* submit a link to your repo via Courseworks

Your solutions to Problems 1, 2, and 3 should be implemented in your .Rmd file, and your git commit history should reflect the process you used to solve these Problems.

For this Problem, we will assess adherence to the instructions above regarding repo structure, git commit history, and whether we are able to knit your .Rmd to ensure that your work is reproducible. Adherence to appropriate styling and clarity of code will be assessed in Problems 1+.


## **Problem 1**

This problem uses the Mr. Trash Wheel dataset, available as an Excel file on the course website.

**Read and clean** the Mr. Trash Wheel sheet:

* specify the sheet in the Excel file and to omit non-data entries (rows with notes / figures; columns containing notes) using arguments in **read_excel**
* use reasonable variable names
* **omit rows** that do not include dumpster-specific data
* **round** the number of sports balls to the nearest integer and **converts** the result to an integer variable (using as.integer)

```{r read and clean Mr. Trash Wheel sheet}

# load the libraries that are necessary
library(tidyverse)
library(readxl)

# to ensure reproductivity
set.seed(1)

trash_wheel_data = 
  # read the data, choose the sheet Mr. Trash Wheel, read only non-figure part, and get rid of notes as a column
  read_excel("./data/HealthyHarborWaterWheelTotals2018-7-28.xlsx", sheet = "Mr. Trash Wheel", range = "A2:N338") %>% 
  # clean names
  janitor::clean_names() %>% 
  # omit rows that do not include dumpster-specific data
  drop_na(dumpster) %>% 
  # round the number of sports balls to the nearest integer and converts the result to an integer variable (using as.integer)
  mutate(sports_balls = as.integer(round(sports_balls)))

###### Question ######
# Why I can't use round(sports_balls) %>%  directly?
######################

trash_wheel_data

```

**Read and clean** precipitation data for 2017 and 2018. For each, **omit rows** without precipitation data and **add a variable year**. Next, combine precipitation datasets and **convert** month to a character variable (the variable month.name is built into R and should be useful).

```{r Read, clean and combine 2017 and 2018}

Prec_2017 = 
  # read the data in the sheet precipation 2017 and skip the first row
  read_excel("./data/HealthyHarborWaterWheelTotals2018-7-28.xlsx", sheet = "2017 Precipitation", skip = 1) %>% 
  # clean names
  janitor::clean_names() %>% 
  # rename total column to precipitation
  rename(precipitation = total) %>% 
  # omit rows without precipitation data
  drop_na(precipitation) %>% 
  # add a variable year
  mutate(year = "2017") %>% 
  # omit total raw
  drop_na(month)

Prec_2017


Prec_2018 = 
  # read the data in the sheet precipation 2018
  read_excel("./data/HealthyHarborWaterWheelTotals2018-7-28.xlsx", sheet = "2018 Precipitation", skip = 1) %>% 
  # clean names
  janitor::clean_names() %>% 
  # rename total column to precipitation
  rename(precipitation = total) %>% 
  # omit rows without precipitation data
  drop_na(precipitation) %>% 
  # add a variable year
  mutate(year = "2018") %>% 
   # omit total raw
  drop_na(month)

Prec_2018

# Now combine two datasets
Prec_data = 
  full_join(Prec_2017, Prec_2018) %>% 
  # convert month to a character variable
  mutate(month = month.name[month],       # [] reads the index
         # make everything in month to their lower cases
         month = str_to_lower(month))    

Prec_data

```

**Write a paragraph about these data**; you are encouraged to use inline R. Be sure to note the **number of observations** in both resulting datasets, and give examples of **key variables**. For available data, what was the **total precipitation in 2018**? What was **the median number of sports balls in a dumpster in 2017**?


## **Problem 2**

This problem uses the FiveThirtyEight data; these data were gathered to create the interactive graphic on this page. In particular, we’ll use the data in pols-month.csv, unemployment.csv, and snp.csv. Our goal is to merge these into a single data frame using year and month as keys across datasets.

First, **clean the data** in pols-month.csv. Use separate() to **break up the variable** mon into integer variables year, month, and day; **replace** month number with month name; **create a president variable** taking values gop and dem, and **remove** prez_dem and prez_gop; and **remove** the day variable.

```{r read and clean pols-month.csv}

pols_month = 
  # read the dataset
  read_csv("./data/pols-month.csv") %>% 
  # clean names
  janitor::clean_names() %>% 
  # Use separate() to break up the variable mon into integer variables year, month, and day
  separate(mon, into = c("year", "month","day"), sep = "-") %>% 
  # change month format from 01 to 1 and so on
  mutate(month = as.integer(month)) %>% 
  
  ########## TEST ###########
  mutate(year = as.integer(year)) %>% 
  ###########################

  # put the month in order from Janurary to December and years from earlier to later
  arrange(year, month) %>% 
  # replace month number with month name
  mutate(month = month.name[month], 
         # make everything in month to their lower cases
         month = str_to_lower(month)) %>% 
  # create a president variable taking values gop and dem, and remove prez_gop and prez_dem
  pivot_longer(
    c("prez_gop", "prez_dem"),
    names_to = "president",
    names_prefix = "prez_",
    values_to = "value") %>% 
  filter(!(value == "0")) %>%    # remove all observations that is a false
  select(year, month, everything(), -value, -day)   # keep only the variable president taking values gop and dem, put year and month as leading columns, and remove day variable

pols_month

```


Second, **clean the data** in snp.csv using a **similar** process to the above. For consistency across datasets, **arrange according to year and month**, and **organize so that year and month are the leading columns**.

```{r clean and read snp.csv}

snp_data = 
  # read the dataset
  read_csv("./data/snp.csv") %>% 
  # clean names
  janitor::clean_names() %>% 
  # Use separate() to break up the variable mon into integer variables year, month, and day
  separate(date, into = c("month", "day","year"), sep = "/") %>% 
  # change month format from 01 to 1 and so on
  mutate(month = as.integer(month)) %>% 
  
  ############# TEST ##############
  mutate(year = as.integer(year)) %>% 
  #################################

  # put the month in order from Janurary to December and years from earlier to later
  arrange(year, month) %>% 
  # replace month number with month name
  mutate(month = month.name[month], 
         # make everything in month to their lower cases
         month = str_to_lower(month)) %>% 
  # put year and month as leading columns, and remove day variable
  select(year, month, everything(), -day)

snp_data

```


Third, **tidy** the unemployment data so that it can be merged with the previous datasets. This process will involve switching from “wide” to “long” format; **ensuring that key variables have the same name**; and **ensuring that key variables take the same values**.

```{r tidy the unemployment data}

unemployment_data = 
  # read the dataset
  read_csv("./data/unemployment.csv") %>% 
  # clean names
  janitor::clean_names() %>% 
  # switch from "wide" to "long"
  pivot_longer(
    jan:dec,
    names_to = "month",
    values_to = "unemployment") %>% 
  # make the name of each month to their full names
    mutate(month = str_subset(month.name, str_to_sentence(month)),
         # make everything in month to their lower cases
         month = str_to_lower(month)) %>% 
  
  ################ TEST #################
  mutate(year = as.integer(year)) %>% 
  #######################################

  # put year and month as leading columns
  select(year, month, everything())

unemployment_data

```


**Join the datasets** by merging snp into pols, and **merging** unemployment into the result.

```{r join datasets}

merge_data = 
  # merging snp into pols
  left_join(pols_month, snp_data, by = c("year","month")) %>% 
  # merging unemployment into the result
  left_join(., unemployment_data, by = c("year","month"))

merge_data

```


**Write a short paragraph about these datasets**. Explain briefly **what each dataset contained**, and **describe the resulting dataset** (e.g. give the dimension, range of years, and names of key variables).

Note: we could have used a date variable as a key instead of creating year and month keys; doing so would help with some kinds of plotting, and be a more accurate representation of the data. Date formats are tricky, though. For more information check out the lubridate package in the tidyverse.


## **Problem 3**

This problem uses data from NYC Open data on the popularity of baby names, and can be downloaded here.

**Load and tidy** the data. Note that, although these data may seem fairly well formatted initially, the names of a categorical predictor and the case structure of string variables changed over time; you’ll need to address this in your data cleaning. Also, some rows seem duplicated, and these will need to be removed (hint: google something like “dplyr remove duplicate rows” to get started).

```{r load and tidy the data}

baby_names = 
  # read the dataset
  read_csv("./data/Popular_Baby_Names.csv") %>% 
  # clean names
  janitor::clean_names() %>%
  # mutate the ethnicity to same names
  mutate_all(~replace(., ethnicity == "ASIAN AND PACI", "ASIAN AND PACIFIC ISLANDER")) %>% 
  mutate_all(~replace(., ethnicity == "BLACK NON HISP", "BLACK NON HISPANIC")) %>% 
  mutate_all(~replace(., ethnicity == "WHITE NON HISP", "WHITE NON HISPANIC")) %>% 
  # remove duplicate rows
  distinct()

baby_names
  
```


Produce a well-structured, reader-friendly table showing the rank in popularity of the name “Olivia” as a female baby name over time; this should have rows for ethnicities and columns for year. Produce a similar table showing the most popular name among male children over time.

```{r produce well-structed tables}

Olivia_female =
  # keep only Olivia and female
  filter(baby_names, childs_first_name == "Olivia", gender == "FEMALE") %>% 
  pivot_wider(
    # find unique identifer
    id_cols = c("ethnicity", "year_of_birth"),
    # put years into column
    names_from = "year_of_birth",
    # put ranks into table cells
    values_from = "rank"
  )# %>% 
  # show only for Olivia as female baby names
  #select(-gender)

Olivia_female

Popular_name_male = 
  filter(baby_names, rank == 1, gender == "MALE") %>% 
  pivot_wider(
    # find unique identifer
    id_cols = c("ethnicity", "year_of_birth"),
    # put years into column
    names_from = "year_of_birth",
    # put names into table cells
    values_from = "childs_first_name"
  ) 

Popular_name_male

```


Finally, for male, white non-hispanic children born in 2016, produce a scatter plot showing the number of children with a name (y axis) against the rank in popularity of that name (x axis).

