---
title: "p8105_hw2_jt3174"
author: "Jingyi"
date: "9/25/2019"
output: github_document
---

## **Problem 1**

```{r read and clean Mr. Trash Wheel sheet}

# load the libraries that are necessary
library(tidyverse)
library(readxl)

# to ensure reproductivity
set.seed(1)

# read and clean the data
trash_wheel_data = 
  # read the data, choose the sheet Mr. Trash Wheel, read only non-figure part, and get rid of notes as a column using read_excel
  read_excel("./data/HealthyHarborWaterWheelTotals2018-7-28.xlsx", sheet = "Mr. Trash Wheel", range = "A2:N338") %>% 
  # clean names and use reasonable variable names
  janitor::clean_names() %>% 
  # omit rows that do not include dumpster-specific data
  drop_na(dumpster) %>% 
  # round the number of sports balls to the nearest integer and converts the result to an integer variable (using as.integer)
  mutate(sports_balls = as.integer(round(sports_balls)))

trash_wheel_data

```


```{r Read, clean and combine 2017 and 2018}

# read and clean 2017 data
Prec_2017 = 
  # read the data in the sheet precipation 2017 and skip the first row
  read_excel("./data/HealthyHarborWaterWheelTotals2018-7-28.xlsx", sheet = "2017 Precipitation", skip = 1) %>% 
  # clean names
  janitor::clean_names() %>% 
  # rename total column to precipitation
  rename(precipitation = total) %>% 
  # omit rows without precipitation data
  drop_na(precipitation) %>% 
  # add a variable year
  mutate(year = "2017") %>% 
  # omit total raw
  drop_na(month)

Prec_2017

# read and clean 2018 data
Prec_2018 = 
  # read the data in the sheet precipation 2018
  read_excel("./data/HealthyHarborWaterWheelTotals2018-7-28.xlsx", sheet = "2018 Precipitation", skip = 1) %>% 
  # clean names
  janitor::clean_names() %>% 
  # rename total column to precipitation
  rename(precipitation = total) %>% 
  # omit rows without precipitation data
  drop_na(precipitation) %>% 
  # add a variable year
  mutate(year = "2018") %>% 
   # omit total raw
  drop_na(month)

Prec_2018

# Now combine two datasets
Prec_data = 
  full_join(Prec_2017, Prec_2018) %>% 
  # convert month to a character variable
  mutate(month = month.name[month],       # [] reads the index
         # make everything in month to their lower cases
         month = str_to_lower(month))    

Prec_data

# to figure out the sum of 2018 precipitation:
Prec_data %>% 
 group_by(year) %>% 
 summarise(count = sum(precipitation))

```

**Data Description:** 
This whole datasets includes information on the dumpter number, date of collection, amount of total litter and litter type. Additional spreadsheets for Professor Trash wheel and for precipitation amounts are also included in the datasets. In the dataset called Prec_data, there are `r nrow(Prec_data)` observations. One of the most important data in this data set is the precipitation. For the key variables, I think the precipitation is the most meaningful one. Some useful statistics of this variable is the average, which is `r mean(pull(Prec_data, precipitation))`, and the variance, which is `r  var(pull(Prec_data, precipitation))`. For the year of 2018, the total precipitation is 23.5. In the dataset called trash_wheel_data, there are `r nrow(trash_wheel_data)`. For the key variables, I think the weight is the most meaningful one. Some useful statistics of this variable is the average, which is `r mean(pull(trash_wheel_data, weight_tons))`, and the variance, which is `r  var(pull(trash_wheel_data, weight_tons))`. The median number of sports balls in a dumpster in 2017 is `r median(pull(trash_wheel_data, sports_balls))`.


## **Problem 2**

```{r read and clean pols-month.csv}

# read and clean data
pols_month = 
  # read the dataset
  read_csv("./data/pols-month.csv") %>% 
  # clean names
  janitor::clean_names() %>% 
  # Use separate() to break up the variable mon into integer variables year, month, and day
  separate(mon, into = c("year", "month","day"), sep = "-") %>% 
  # change month format from 01 to 1 and so on
  mutate(month = as.integer(month)) %>% 
  mutate(year = as.integer(year)) %>% 
  # put the month in order from Janurary to December and years from earlier to later
  arrange(year, month) %>% 
  # replace month number with month name
  mutate(month = month.name[month], 
         # make everything in month to their lower cases
         month = str_to_lower(month)) %>% 
  # create a president variable taking values gop and dem, and remove prez_gop and prez_dem
  pivot_longer(
    c("prez_gop", "prez_dem"),
    names_to = "president",
    names_prefix = "prez_",
    values_to = "value") %>% 
  filter(!(value == "0")) %>%    # remove all observations that is a false
  select(year, month, everything(), -value, -day)   # keep only the variable president taking values gop and dem, put year and month as leading columns, and remove day variable

pols_month

```


```{r clean and read snp.csv}

# read and clean the data
snp_data = 
  # read the dataset
  read_csv("./data/snp.csv") %>% 
  # clean names
  janitor::clean_names() %>% 
  # Use separate() to break up the variable mon into integer variables year, month, and day
  separate(date, into = c("month", "day","year"), sep = "/") %>% 
  # change month format from 01 to 1 and so on
  mutate(month = as.integer(month)) %>% 
  mutate(year = as.integer(year)) %>% 
  # put the month in order from Janurary to December and years from earlier to later
  arrange(year, month) %>% 
  # replace month number with month name
  mutate(month = month.name[month], 
         # make everything in month to their lower cases
         month = str_to_lower(month)) %>% 
  # put year and month as leading columns, and remove day variable
  select(year, month, everything(), -day)

snp_data

```


```{r tidy the unemployment data}

# read, clean and tidy the unemployment data
unemployment_data = 
  # read the dataset
  read_csv("./data/unemployment.csv") %>% 
  # clean names
  janitor::clean_names() %>% 
  # switch from "wide" to "long" format
  pivot_longer(
    jan:dec,
    names_to = "month",
    values_to = "unemployment") %>% 
  # make the name of each month to their full names (aka. same names)
    mutate(month = str_subset(month.name, str_to_sentence(month)),
         # make everything in month to their lower cases
         month = str_to_lower(month)) %>% 
  mutate(year = as.integer(year)) %>% 
  # put year and month as leading columns
  select(year, month, everything())

unemployment_data

```


```{r join datasets}

merge_data = 
  # merging snp into pols
  left_join(pols_month, snp_data, by = c("year","month")) %>% 
  # merging unemployment into the result
  left_join(., unemployment_data, by = c("year","month"))

merge_data

```


**Data Description:** 
In the dataset called pols_month, there are `r nrow(pols_month)` observations of `r ncol(pols_month)` variables related to the number of national politicians who are democratic or republic at any given time. There are date varaibles, number of republican or democratic governors/senators/representatives/president on the associated date. The range of dates is from 1949 to 2015. Some of the key variables are president and year. 
In the dataset called snp_data, there are `r nrow(snp_data)` observations of `r ncol(snp_data)` variables related to Standard & Poor’s stock market index (S&P), often used as a representative measure of stock market as a whole. There are year, month and close values of the S&P stock index on the associated date. The range of dates is from 1950 to 2015. Some of the key variables are close, year and month. 
In the dataset called unemployment_data, there are `r nrow(unemployment_data)` observations of `r ncol(unemployment_data)` variables related to the precentage of unemployment on the associated dates, which is pretty self-explanatory. The range of dates is from 
1948 to 2015. Some of the key variables are umemployment, year and month. 
In the dataset called merge_data, this is pretty much the same with what was stated above, because this dataset is just a combination with all three datasets above. There are `r nrow(merge_data)` observations of `r ncol(merge_data)` variables.  

## **Problem 3**


```{r load and tidy the data}

# load and tidy the data
baby_names = 
  # read the dataset
  read_csv("./data/Popular_Baby_Names.csv", col_types = "icccii") %>% 
  # clean names
  janitor::clean_names() %>%
  # mutate the all different observations in ethnicity to same names
  mutate(ethnicity = replace(ethnicity, ethnicity == "ASIAN AND PACI", "ASIAN AND PACIFIC ISLANDER")) %>% 
  mutate(ethnicity = replace(ethnicity, ethnicity == "BLACK NON HISP", "BLACK NON HISPANIC")) %>% 
  mutate(ethnicity = replace(ethnicity, ethnicity == "WHITE NON HISP", "WHITE NON HISPANIC")) %>% 
  # change to lower case
  mutate(childs_first_name = str_to_lower(childs_first_name)) %>% 
  # remove duplicate rows
  distinct()

baby_names
  
```



```{r produce well-structed tables}

# Make a dataframe for the table first for Olivia as female baby names
Olivia_female =
  # keep only Olivia and female
  filter(baby_names, childs_first_name == "Olivia", gender == "FEMALE") %>% 
  pivot_wider(
    # find unique identifer
    id_cols = c("ethnicity", "year_of_birth"),
    # put years into column
    names_from = "year_of_birth",
    # put ranks into table cells
    values_from = "rank"
  )

Olivia_female

# make a dataframe for the male's most popular names
Popular_name_male = 
  # keep only male babies and the top-ranked name
  filter(baby_names, rank == 1, gender == "MALE") %>% 
  pivot_wider(
    # find unique identifer
    id_cols = c("ethnicity", "year_of_birth"),
    # put years into column
    names_from = "year_of_birth",
    # put names into table cells
    values_from = "childs_first_name"
  ) 

Popular_name_male

```


```{r scatterplot}

# make a dataframe for male, white non-hispanic children born in 2016 with number of children with a name and their rank in popularity of that name
white_non_hisp_2016 = 
  # keep only 2016-borned, male, white non-hispanic babies
  filter(baby_names, year_of_birth == 2016, gender == "MALE", ethnicity == "WHITE NON HISPANIC") %>% 
  # convert count and rank to integers
  mutate(count = as.integer(count), rank = as.integer(rank))

white_non_hisp_2016

# make the scatterplot to have x-axis as rank and y-axis as name for this dataframe
ggplot(data = white_non_hisp_2016) + geom_point(aes(x = rank, y = count))


```

